---
title: "Obesity and Healthcare Spending Among Medicare Beneficiaries"
subtitle: "MEPS 2020 & 2022 | Naive GLM vs. Instrumental Variable Analysis"
date: today
format:
  pdf:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 4.5
    df-print: kable
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
execute:
  warning: false
  message: false
  cache: true
---

```{r setup}
#| cache: false
library(tidyverse)
library(tidymodels)
library(ggthemes)
library(skimr)
library(srvyr)
library(gtsummary)
library(scales)

source("functions.R")

wsj_navy <- "#0F3F6D"
wsj_red  <- "#C41E3D"
```

---

## Overview

This report analyzes the relationship between obesity and healthcare spending among Medicare
beneficiaries using the Medical Expenditure Panel Survey (MEPS). Two distinct analytical
approaches are compared:

1. **My Analysis** — A survey-weighted Gamma GLM using categorical BMI class as the primary
   predictor, estimating spending differences across BMI groups and projecting savings from
   stepping beneficiaries down one obesity class.

2. **Article Replication** — A 2-Stage Residual Inclusion (2SRI) instrumental variable model
   following Cawley et al., using the BMI of the oldest household child as an instrument for
   the parent's BMI to address endogeneity. Savings are projected at 5–25% continuous BMI
   reductions.

---

## Data

### Source

Data come from the **Medical Expenditure Panel Survey Household Component (MEPS-HC)**, a
nationally representative survey of the US civilian non-institutionalized population. MEPS
collects longitudinal data on health conditions, medical expenditures, and health insurance.

**Survey years used: 2020 and 2022.** BMI (ADBMI42) is collected only in even-numbered survey
years (from 2018 onward), limiting the pooled sample to these two rounds.

**Files used:**

| File | Year | Description |
|------|------|-------------|
| h222 / h241 | 2020 / 2022 | Conditions files — CCSR-coded diagnoses |
| h224 / h243 | 2020 / 2022 | Consolidated person-level files — demographics, expenditures, weights |

### Target Conditions

The analysis focuses on Medicare beneficiaries with at least one of **10 target clinical
conditions** identified via CCSR codes:

- Diabetes, Hyperlipidemia, Hypertension
- Mental health conditions, Pulmonary disease, Asthma
- Arthritis, Back problems, Heart disease, Cerebrovascular disease

---

## Data Pipeline

### Conditions Data

```{r load-conditions}
#| cache: true
target_ccsr <- c(
    "END002","END003","END004","END005","END006",
    "END010",
    "CIR007","CIR008",
    "MBD001","MBD002","MBD003","MBD004","MBD005","MBD006","MBD007","MBD008",
    "MBD009","MBD010","MBD011","MBD012","MBD013","MBD014","MBD017","MBD018",
    "MBD019","MBD020","MBD021","MBD022","MBD023","MBD024","MBD025","MBD026",
    "MBD027","MBD028","MBD029","MBD030","MBD031","MBD032","MBD033","MBD034",
    "RSP006","RSP007","RSP008","RSP010","RSP011","RSP012","RSP013","RSP014","RSP016",
    "MUS001","MUS002","MUS003","MUS004","MUS005","MUS006","MUS007",
    "MUS011","MUS038",
    "CIR001","CIR002","CIR003","CIR004","CIR005","CIR006","CIR010","CIR011",
    "CIR012","CIR014","CIR015","CIR016","CIR017","CIR018",
    "NVS012","CIR020","CIR021","CIR022","CIR023","CIR024","CIR025",
    "RSP009"
)

h222 <- load_meps_conditions(
    dat_path    = "data/conditions_data/2020/h222.dat",
    spec_path   = "setup/h222_spec.csv",
    rename_path = "setup/h222_renames.csv",
    target_ccsr = target_ccsr,
    ccsr_cols   = c("ccsr_code1","ccsr_code2","ccsr_code3"),
    survey_year = 2020
)

h241 <- load_meps_conditions(
    dat_path    = "data/conditions_data/2022/h241.dat",
    spec_path   = "setup/h241_spec.csv",
    rename_path = "setup/h241_renames.csv",
    target_ccsr = target_ccsr,
    ccsr_cols   = c("ccsr_code1","ccsr_code2","ccsr_code3","ccsr_code4"),
    survey_year = 2022
)

conditions_stacked <- bind_rows(h222, h241)
rm(h222, h241); gc()
```

### Consolidated Data

```{r load-consolidated}
#| cache: true
consolidated_cols_shared <- c(
    panel_number = "panel_number",
    family_id    = "annual_family_identifier",
    sex          = "sex",
    race         = "race_edited_imputed_racev1x",
    hispanic     = "hispanic_ethnicity_edited_imputed_hispanx",
    education    = "years_of_educ_when_first_entered_meps",
    bmi          = "adult_body_mass_index_17_rd_4_2",
    child_bmi    = "child_s_body_mass_index_6_17_r4_2"
)

consolidated_cols_2020 <- c(
    person_id        = "person_id_duid_pid",
    consolidated_cols_shared,
    age              = "age_as_of_12_31_20_edited_imputed",
    region           = "census_region_as_of_12_31_20",
    marital_status   = "marital_status_12_31_20_edited_imputed",
    ever_medicare    = "ever_have_medicare_during_2020_ed",
    mcr_jan          = "covered_by_medicare_in_jan20_ed",
    mcr_feb          = "covered_by_medicare_in_feb20_ed",
    mcr_mar          = "covered_by_medicare_in_mar20_ed",
    mcr_apr          = "covered_by_medicare_in_apr20_ed",
    mcr_may          = "covered_by_medicare_in_may20_ed",
    mcr_jun          = "covered_by_medicare_in_jun20_ed",
    mcr_jul          = "covered_by_medicare_in_jul20_ed",
    mcr_aug          = "covered_by_medicare_in_aug20_ed",
    mcr_sep          = "covered_by_medicare_in_sep20_ed",
    mcr_oct          = "covered_by_medicare_in_oct20_ed",
    mcr_nov          = "covered_by_medicare_in_nov20_ed",
    mcr_dec          = "covered_by_medicare_in_dec20_ed",
    total_expenditure = "total_health_care_exp_20",
    person_weight    = "final_person_weight_2020",
    saq_weight       = "final_saq_person_weight_2020",
    variance_stratum = "variance_estimation_stratum_2020",
    variance_psu     = "variance_estimation_psu_2020"
)

consolidated_cols_2022 <- c(
    person_id        = "person_id_duid_pid",
    consolidated_cols_shared,
    age              = "age_as_of_12_31_22_edited_imputed",
    region           = "census_region_as_of_12_31_22",
    marital_status   = "marital_status_12_31_22_edited_imputed",
    ever_medicare    = "ever_have_medicare_during_2022_ed",
    mcr_jan          = "covered_by_medicare_in_jan22_ed",
    mcr_feb          = "covered_by_medicare_in_feb22_ed",
    mcr_mar          = "covered_by_medicare_in_mar22_ed",
    mcr_apr          = "covered_by_medicare_in_apr22_ed",
    mcr_may          = "covered_by_medicare_in_may22_ed",
    mcr_jun          = "covered_by_medicare_in_jun22_ed",
    mcr_jul          = "covered_by_medicare_in_jul22_ed",
    mcr_aug          = "covered_by_medicare_in_aug22_ed",
    mcr_sep          = "covered_by_medicare_in_sep22_ed",
    mcr_oct          = "covered_by_medicare_in_oct22_ed",
    mcr_nov          = "covered_by_medicare_in_nov22_ed",
    mcr_dec          = "covered_by_medicare_in_dec22_ed",
    total_expenditure = "total_health_care_exp_22",
    person_weight    = "final_person_weight_2022",
    saq_weight       = "final_saq_person_weight_2022",
    variance_stratum = "variance_estimation_stratum_2022",
    variance_psu     = "variance_estimation_psu_2022"
)

h224 <- load_meps_consolidated(
    dat_path    = "data/consolidated_data/2020/H224.DAT",
    spec_path   = "setup/h224_spec.csv",
    rename_path = "setup/h224_renames.csv",
    survey_year = 2020,
    col_map     = consolidated_cols_2020
)

h243 <- load_meps_consolidated(
    dat_path    = "data/consolidated_data/2022/h243.dat",
    spec_path   = "setup/h243_spec.csv",
    rename_path = "setup/h243_renames.csv",
    survey_year = 2022,
    col_map     = consolidated_cols_2022
)

consolidated_stacked <- bind_rows(h224, h243)
rm(h224, h243); gc()
```

### Sample Construction & Cleaning

```{r build-data-clean}
#| cache: true
ccsr_to_condition <- c(
    END002="diabetes", END003="diabetes", END004="diabetes", END005="diabetes", END006="diabetes",
    END010="hyperlipidemia",
    CIR007="hypertension", CIR008="hypertension",
    MBD001="mental_health", MBD002="mental_health", MBD003="mental_health", MBD004="mental_health",
    MBD005="mental_health", MBD006="mental_health", MBD007="mental_health", MBD008="mental_health",
    MBD009="mental_health", MBD010="mental_health", MBD011="mental_health", MBD012="mental_health",
    MBD013="mental_health", MBD014="mental_health", MBD017="mental_health", MBD018="mental_health",
    MBD019="mental_health", MBD020="mental_health", MBD021="mental_health", MBD022="mental_health",
    MBD023="mental_health", MBD024="mental_health", MBD025="mental_health", MBD026="mental_health",
    MBD027="mental_health", MBD028="mental_health", MBD029="mental_health", MBD030="mental_health",
    MBD031="mental_health", MBD032="mental_health", MBD033="mental_health", MBD034="mental_health",
    RSP006="pulmonary_disease", RSP007="pulmonary_disease", RSP008="pulmonary_disease",
    RSP010="pulmonary_disease", RSP011="pulmonary_disease", RSP012="pulmonary_disease",
    RSP013="pulmonary_disease", RSP014="pulmonary_disease", RSP016="pulmonary_disease",
    MUS001="arthritis", MUS002="arthritis", MUS003="arthritis", MUS004="arthritis",
    MUS005="arthritis", MUS006="arthritis", MUS007="arthritis",
    MUS011="back_problems", MUS038="back_problems",
    CIR001="heart_disease", CIR002="heart_disease", CIR003="heart_disease", CIR004="heart_disease",
    CIR005="heart_disease", CIR006="heart_disease", CIR010="heart_disease", CIR011="heart_disease",
    CIR012="heart_disease", CIR014="heart_disease", CIR015="heart_disease", CIR016="heart_disease",
    CIR017="heart_disease", CIR018="heart_disease",
    NVS012="cerebrovascular_disease", CIR020="cerebrovascular_disease",
    CIR021="cerebrovascular_disease", CIR022="cerebrovascular_disease",
    CIR023="cerebrovascular_disease", CIR024="cerebrovascular_disease",
    CIR025="cerebrovascular_disease",
    RSP009="asthma"
)

mcr_months <- paste0("mcr_", c("jan","feb","mar","apr","may","jun",
                                "jul","aug","sep","oct","nov","dec"))

condition_vars <- c("diabetes","hypertension","hyperlipidemia","mental_health",
                    "pulmonary_disease","arthritis","back_problems","heart_disease",
                    "cerebrovascular_disease","asthma")

data_joined <- consolidated_stacked |>
    inner_join(conditions_stacked, by = c("person_id","panel_number","survey_year")) |>
    mutate(across(starts_with("ccsr_code"), ~ ccsr_to_condition[.]))

data_medicare <- data_joined |>
    filter(if_all(all_of(mcr_months), ~ . == 1))

person_cols <- c("person_id","panel_number","family_id","sex","race","hispanic",
                 "education","bmi","age","region","marital_status","ever_medicare",
                 mcr_months,"total_expenditure","person_weight","saq_weight",
                 "variance_stratum","variance_psu","survey_year")

condition_flags <- data_medicare |>
    select(person_id, survey_year, starts_with("ccsr_code")) |>
    pivot_longer(starts_with("ccsr_code"), values_to = "condition", names_to = NULL) |>
    filter(!is.na(condition)) |>
    distinct() |>
    mutate(present = 1L) |>
    pivot_wider(names_from = condition, values_from = present, values_fill = 0L)

data_clean <- data_medicare |>
    select(all_of(person_cols)) |>
    distinct(person_id, survey_year, .keep_all = TRUE) |>
    filter(!is.na(bmi), between(bmi, 10, 100)) |>
    filter(!is.na(total_expenditure), total_expenditure > 0) |>
    filter(!is.na(person_weight), person_weight > 0) |>
    filter(!is.na(sex), !is.na(race), !is.na(hispanic),
           !is.na(education), !is.na(marital_status), !is.na(region)) |>
    select(-all_of(mcr_months), -ever_medicare) |>
    inner_join(condition_flags, by = c("person_id","survey_year")) |>
    mutate(
        bmi_category = factor(
            case_when(
                bmi < 25 ~ "normal_or_below",
                bmi < 30 ~ "overweight",
                bmi < 35 ~ "obese_class_i",
                bmi < 40 ~ "obese_class_ii",
                bmi >= 40 ~ "obese_class_iii"
            ),
            levels = c("normal_or_below","overweight","obese_class_i",
                       "obese_class_ii","obese_class_iii")
        )
    ) |>
    mutate(across(any_of(condition_vars), as.logical)) |>
    mutate(
        sex = factor(sex, levels = c(1,2), labels = c("Male","Female")),
        race = factor(race, levels = 1:6,
                      labels = c("White","Black","Am. Indian / Alaska Native",
                                 "Asian","Nat. Hawaiian / Pacific Islander","Multiple races")),
        hispanic = factor(hispanic, levels = c(1,2), labels = c("Hispanic","Not Hispanic")),
        marital_status = factor(marital_status, levels = 1:5,
                                labels = c("Married","Widowed","Divorced","Separated","Never married")),
        region = factor(region, levels = 1:4,
                        labels = c("Northeast","Midwest","South","West"))
    )

rm(data_joined, data_medicare, condition_flags, consolidated_stacked, conditions_stacked)
invisible(gc())
```

The cleaning pipeline applies the following sequential filters:

| Step | Filter | Rationale |
|------|--------|-----------|
| 1 | Inner join to conditions | Keep only persons with ≥ 1 target condition |
| 2 | 12-month Medicare coverage | Continuous enrollment ensures complete spending capture |
| 3 | Adult BMI 10–100 | Implausible values removed; children excluded |
| 4 | Expenditure > 0 | Gamma family requires strictly positive outcomes |
| 5 | Positive survey weight | Persons with missing/zero weights excluded |
| 6 | Complete demographics | Listwise deletion on sex, race, Hispanic, education, marital status, region |
| 7 | BMI categorization | Underweight collapsed into normal due to sparse counts in Medicare population |
| 8 | Factor recoding | All categorical variables labeled for interpretable output |

```{r sample-size}
#| echo: false
cat("Final analytic sample:", nrow(data_clean), "person-years\n")
cat("  2020:", sum(data_clean$survey_year == 2020), "\n")
cat("  2022:", sum(data_clean$survey_year == 2022), "\n")
```

### Survey Design

All estimates account for MEPS's stratified clustered design.

```{r survey-design}
svy_clean <- data_clean |>
    as_survey_design(
        ids     = variance_psu,
        strata  = variance_stratum,
        weights = person_weight,
        nest    = TRUE
    )
```

---

## Descriptive Statistics

### Table 1: Sample Characteristics by BMI Category

```{r table1}
#| cache: true
tbl_svysummary(
    svy_clean,
    by      = bmi_category,
    include = c(age, sex, race, hispanic, education, marital_status, region,
                total_expenditure, all_of(condition_vars)),
    statistic = list(all_continuous() ~ "{mean} ({sd})",
                     all_categorical() ~ "{p}% ({n})"),
    digits    = list(all_continuous() ~ 1, all_categorical() ~ c(1, 0)),
    missing   = "no",
    label     = list(
        age ~ "Age (years)", sex ~ "Sex", race ~ "Race",
        hispanic ~ "Hispanic ethnicity", education ~ "Education (years)",
        marital_status ~ "Marital status", region ~ "Census region",
        total_expenditure ~ "Total expenditure ($)",
        diabetes ~ "Diabetes", hypertension ~ "Hypertension",
        hyperlipidemia ~ "Hyperlipidemia",
        mental_health ~ "Mental health condition",
        pulmonary_disease ~ "Pulmonary disease", arthritis ~ "Arthritis",
        back_problems ~ "Back problems", heart_disease ~ "Heart disease",
        cerebrovascular_disease ~ "Cerebrovascular disease", asthma ~ "Asthma"
    )
) |>
    add_p(test = list(all_continuous() ~ "svy.kruskal.test",
                      all_categorical() ~ "svy.chisq.test")) |>
    add_overall() |>
    bold_labels() |>
    bold_p(t = 0.05)
```

### Condition Prevalence Heatmap

```{r heatmap}
#| fig-height: 6
heatmap_data <- condition_vars |>
    map(\(cond) {
        svy_clean |>
            group_by(bmi_category) |>
            summarise(prevalence = survey_mean(as.numeric(.data[[cond]]),
                                               na.rm = TRUE) * 100) |>
            mutate(condition = cond)
    }) |>
    list_rbind() |>
    mutate(condition = str_replace_all(condition, "_", " ") |> str_to_title())

bmi_labels <- c(normal_or_below = "Normal/Below", overweight = "Overweight",
                obese_class_i = "Obese I", obese_class_ii = "Obese II",
                obese_class_iii = "Obese III")

ggplot(heatmap_data, aes(x = bmi_category, y = condition, fill = prevalence)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.1f%%", prevalence)), size = 3, color = "black") +
    scale_fill_gradient(low = "#f7fbff", high = "#2166ac", name = "Prevalence (%)") +
    scale_x_discrete(labels = bmi_labels) +
    labs(title = "Condition Prevalence by BMI Category",
         subtitle = "Medicare beneficiaries, 2020 & 2022 | Survey-weighted estimates",
         x = NULL, y = NULL) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1),
          panel.grid = element_blank(),
          plot.title = element_text(face = "bold"))
```

---

## My Analysis: Naive Gamma GLM

### Approach

This analysis treats BMI class as an observed categorical predictor and estimates its
association with total healthcare expenditure using a **survey-weighted Gamma GLM with a
log link** — the standard family for non-negative, right-skewed cost data. BMI category
is the primary predictor; covariates include age, sex, race, Hispanic ethnicity, education,
marital status, region, survey year, and all 10 condition flags.

Spending differences are summarized as **exponentiated coefficients (spending ratios)**
relative to the Normal/Underweight reference group. Savings are then projected by
counterfactually assigning each obese person to the next lower BMI class, holding all
other covariates fixed.

### Spending Gradient

```{r spending-gradient}
spend_by_bmi <- svy_clean |>
    group_by(bmi_category) |>
    summarise(mean = survey_mean(total_expenditure, vartype = "ci", na.rm = TRUE))

ggplot(spend_by_bmi, aes(x = bmi_category, y = mean)) +
    geom_col(fill = wsj_navy, width = 0.65) +
    geom_errorbar(aes(ymin = mean_low, ymax = mean_upp),
                  width = 0.2, linewidth = 0.7, color = "gray40") +
    geom_text(aes(label = dollar(round(mean, -2))),
              vjust = -0.9, size = 3.5, fontface = "bold") +
    scale_x_discrete(labels = c(normal_or_below = "Normal /\nBelow",
                                 overweight = "Overweight",
                                 obese_class_i = "Obese\nClass I",
                                 obese_class_ii = "Obese\nClass II",
                                 obese_class_iii = "Obese\nClass III")) +
    scale_y_continuous(labels = dollar_format(), expand = expansion(mult = c(0, 0.18))) +
    labs(title = "Healthcare Spending Rises With Obesity",
         subtitle = "Survey-weighted mean annual expenditure | Medicare, 2020 & 2022",
         x = NULL, y = "Mean annual expenditure") +
    theme_wsj(base_size = 11) +
    theme(axis.title.y = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

### Adjusted Spending Ratios

```{r model-spend}
#| cache: true
model_spend <- survey::svyglm(
    total_expenditure ~ bmi_category + age + sex + race + hispanic +
        education + marital_status + region + as.factor(survey_year) +
        diabetes + hypertension + hyperlipidemia + mental_health +
        pulmonary_disease + arthritis + back_problems + heart_disease +
        cerebrovascular_disease + asthma,
    design = svy_clean,
    family = Gamma(link = "log")
)
```

```{r plot-spend-coefs}
bmi_coefs <- tibble(
    term      = names(coef(model_spend)),
    estimate  = coef(model_spend),
    conf.low  = confint(model_spend)[, 1],
    conf.high = confint(model_spend)[, 2]
) |>
    filter(str_detect(term, "bmi_category")) |>
    mutate(
        across(c(estimate, conf.low, conf.high), exp),
        label = str_remove(term, "bmi_category") |>
            str_replace_all("_", " ") |> str_to_title()
    )

ggplot(bmi_coefs, aes(x = estimate, y = fct_reorder(label, estimate))) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.7) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.2, linewidth = 0.8, color = "gray40") +
    geom_point(size = 4, color = wsj_red) +
    geom_text(aes(label = sprintf("×%.2f", estimate)),
              hjust = -0.35, size = 3.5, fontface = "bold") +
    scale_x_continuous(limits = c(0.9, max(bmi_coefs$conf.high) * 1.18),
                       labels = \(x) sprintf("×%.1f", x)) +
    labs(title = "Obesity Linked to Higher Spending After Adjustment",
         subtitle = "Spending ratio vs. Normal/Underweight | Gamma GLM, survey-weighted",
         x = "Adjusted spending ratio (reference: Normal / Underweight)", y = NULL) +
    theme_wsj(base_size = 11) +
    theme(axis.title.x = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

### Counterfactual Savings Projection

Each obese beneficiary is counterfactually assigned to the next lower BMI category
(Class III → II, Class II → I, Class I → Overweight), with all other covariates held
fixed. The predicted spending difference is the estimated per-person annual saving.

```{r counterfactual}
#| cache: true
bmi_levels <- levels(data_clean$bmi_category)

data_cf <- data_clean |>
    mutate(bmi_category = factor(
        case_when(
            bmi_category == "obese_class_iii" ~ "obese_class_ii",
            bmi_category == "obese_class_ii"  ~ "obese_class_i",
            bmi_category == "obese_class_i"   ~ "overweight",
            TRUE ~ as.character(bmi_category)
        ), levels = bmi_levels
    ))

pred_orig <- as.numeric(predict(model_spend, newdata = data_clean, type = "response"))
pred_cf   <- as.numeric(predict(model_spend, newdata = data_cf,    type = "response"))

savings_data <- data_clean |>
    mutate(spend_pred = pred_orig, spend_cf = pred_cf,
           spend_saving = spend_pred - spend_cf) |>
    filter(bmi_category %in% c("obese_class_i","obese_class_ii","obese_class_iii")) |>
    group_by(bmi_category) |>
    summarise(mean_saving_pp = weighted.mean(spend_saving, w = person_weight),
              total_saving   = sum(spend_saving * person_weight),
              .groups = "drop")

savings_labels <- c(obese_class_i   = "Obese Class I\n→ Overweight",
                    obese_class_ii  = "Obese Class II\n→ Class I",
                    obese_class_iii = "Obese Class III\n→ Class II")
```

```{r plot-savings-naive}
#| layout-ncol: 2
ggplot(savings_data, aes(x = bmi_category, y = mean_saving_pp)) +
    geom_col(fill = wsj_red, width = 0.6) +
    geom_text(aes(label = dollar(round(mean_saving_pp, -1))),
              vjust = -0.6, size = 4, fontface = "bold") +
    scale_x_discrete(labels = savings_labels) +
    scale_y_continuous(labels = dollar_format(), expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Predicted Annual Savings Per Person",
         subtitle = "One BMI class reduction | Naive GLM",
         x = NULL, y = "Mean savings per person") +
    theme_wsj(base_size = 11) +
    theme(axis.title.y = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))

ggplot(savings_data, aes(x = bmi_category, y = total_saving / 1e9)) +
    geom_col(fill = wsj_navy, width = 0.6) +
    geom_text(aes(label = sprintf("$%.1fB", total_saving / 1e9)),
              vjust = -0.6, size = 4, fontface = "bold") +
    scale_x_discrete(labels = savings_labels) +
    scale_y_continuous(labels = \(x) sprintf("$%.0fB", x),
                       expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Estimated Total Medicare Savings",
         subtitle = "Population-weighted | Naive GLM",
         x = NULL, y = "Total savings (billions)") +
    theme_wsj(base_size = 11) +
    theme(axis.title.y = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

---

## Article Replication: 2SRI Instrumental Variable Model

### Approach

This replication follows Cawley et al. and addresses a fundamental limitation of the
naive model: **BMI is endogenous**. Unobserved factors (genetics, lifestyle, socioeconomic
circumstances) simultaneously influence both BMI and healthcare utilization, meaning the
naive GLM coefficient on BMI conflates the causal effect of adiposity with these
confounders.

The solution is a **2-Stage Residual Inclusion (2SRI)** instrumental variable estimator:

- **Stage 1:** Regress BMI on the instrument and all covariates. The instrument is the
  BMI of the oldest household child aged 11–45, identified through MEPS family linkages.
  Child BMI is a strong predictor of parent BMI — up to 75% of variation in weight is
  genetic — but is plausibly unrelated to the parent's medical spending after conditioning
  on estimated BMI and covariates.

- **Stage 2:** Estimate the Gamma GLM for expenditures including both BMI and the
  Stage 1 residuals. The residual term absorbs the endogenous component of BMI
  (Terza et al. 2008), producing a consistent IV estimate.

### Instrument Construction

```{r build-family-data}
#| cache: true
family_cols_2020 <- c(
    person_id = "person_id_duid_pid", family_id = "annual_family_identifier",
    panel_number = "panel_number", age = "age_as_of_12_31_20_edited_imputed",
    bmi_adult = "adult_body_mass_index_17_rd_4_2",
    bmi_child = "child_s_body_mass_index_6_17_r4_2",
    family_size = "ru_size_including_student_as_of_12_31_20",
    proxy = "was_respondent_a_proxy_in_r4_2",
    dobmm = "date_of_birth_month", dobyy = "date_of_birth_year"
)
family_cols_2022 <- c(
    person_id = "person_id_duid_pid", family_id = "annual_family_identifier",
    panel_number = "panel_number", age = "age_as_of_12_31_22_edited_imputed",
    bmi_adult = "adult_body_mass_index_17_rd_4_2",
    bmi_child = "child_s_body_mass_index_6_17_r4_2",
    family_size = "ru_size_including_student_as_of_12_31_22",
    proxy = "was_respondent_a_proxy_in_r4_2",
    dobmm = "date_of_birth_month", dobyy = "date_of_birth_year"
)

family_all <- bind_rows(
    load_meps_consolidated("data/consolidated_data/2020/H224.DAT",
                           "setup/h224_spec.csv", "setup/h224_renames.csv",
                           2020, family_cols_2020),
    load_meps_consolidated("data/consolidated_data/2022/h243.dat",
                           "setup/h243_spec.csv", "setup/h243_renames.csv",
                           2022, family_cols_2022)
)

instrument_candidates <- family_all |>
    mutate(
        bmi_combined = case_when(
            age >= 11 & age <= 17 & !is.na(bmi_child) ~ bmi_child,
            age >= 18 & age <= 45 & !is.na(bmi_adult) ~ bmi_adult,
            TRUE ~ NA_real_
        ),
        child_age_months = (survey_year - dobyy) * 12 - (dobmm - 12)
    ) |>
    filter(!is.na(bmi_combined), age >= 11, age <= 45, !is.na(child_age_months)) |>
    select(family_id, panel_number, survey_year,
           child_person_id = person_id, child_bmi_instrument = bmi_combined,
           child_age_months, child_age = age)

instrument_table <- data_clean |>
    select(person_id, family_id, panel_number, survey_year) |>
    inner_join(instrument_candidates, by = c("family_id","panel_number","survey_year")) |>
    filter(person_id != child_person_id) |>
    group_by(person_id, survey_year) |>
    slice_max(child_age, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(person_id, survey_year, child_bmi_instrument, child_age_months)

parent_extra <- family_all |>
    semi_join(data_clean, by = c("person_id","survey_year")) |>
    select(person_id, survey_year, family_size, proxy)

rm(family_all); gc()

data_iv <- data_clean |>
    inner_join(instrument_table, by = c("person_id","survey_year")) |>
    left_join(parent_extra,      by = c("person_id","survey_year")) |>
    filter(!is.na(family_size), !is.na(proxy)) |>
    mutate(
        race_eth = factor(
            case_when(
                hispanic == "Hispanic"                       ~ "Hispanic",
                race == "White" & hispanic == "Not Hispanic" ~ "Non-Hispanic White",
                race == "Black" & hispanic == "Not Hispanic" ~ "Non-Hispanic Black",
                race == "Asian"                              ~ "Asian",
                TRUE                                         ~ "Other"
            ),
            levels = c("Non-Hispanic White","Non-Hispanic Black","Hispanic","Asian","Other")
        ),
        age_group = cut(age, breaks = c(0,64,69,74,79,84,Inf),
                        labels = c("<65","65-69","70-74","75-79","80-84","85+"),
                        right = TRUE),
        proxy_reported = factor(proxy, levels = c(1,2), labels = c("Proxy","Self"))
    )
```

```{r iv-sample-size}
#| echo: false
cat("IV analytic sample:", nrow(data_iv), "person-years\n")
cat("  (", round(nrow(data_iv)/nrow(data_clean)*100, 1),
    "% of data_clean — adults with qualifying household child)\n")
```

The IV sample is a subset of the main analytic sample, restricted to Medicare
beneficiaries who have a household child aged 11–45 with a valid BMI recorded in the
same survey year. This sample reduction is expected and is acknowledged as a limitation.

### Instrument Strength

```{r instrument-scatter}
ggplot(data_iv |> slice_sample(n = min(3000, nrow(data_iv))),
       aes(x = child_bmi_instrument, y = bmi)) +
    geom_jitter(alpha = 0.25, size = 0.9, color = wsj_navy, width = 0.3, height = 0.3) +
    geom_smooth(method = "lm", color = wsj_red, linewidth = 1.2, se = TRUE) +
    labs(title = "Child BMI Predicts Parent BMI",
         subtitle = "Instrument validity check | Medicare adults with household child",
         x = "Oldest child BMI (instrument)", y = "Parent BMI") +
    theme_wsj(base_size = 11) +
    theme(axis.title = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

### Stage 1: First-Stage Regression

```{r stage1}
#| cache: true
svy_iv <- data_iv |>
    as_survey_design(ids = variance_psu, strata = variance_stratum,
                     weights = person_weight, nest = TRUE)

stage1_model <- survey::svyglm(
    bmi ~ child_bmi_instrument + age_group + sex + race_eth + education +
        marital_status + region + family_size + proxy_reported +
        child_age_months + as.factor(survey_year),
    design = svy_iv, family = gaussian()
)

stage1_ftest <- survey::regTermTest(stage1_model, ~ child_bmi_instrument)
```

```{r stage1-plot}
stage1_coef <- tibble(
    term      = names(coef(stage1_model)),
    estimate  = coef(stage1_model),
    conf.low  = confint(stage1_model)[, 1],
    conf.high = confint(stage1_model)[, 2]
) |> filter(term == "child_bmi_instrument")

ggplot(stage1_coef, aes(x = estimate, y = "Child BMI -> Parent BMI")) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.7) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.15, linewidth = 1, color = "gray40") +
    geom_point(size = 5, color = wsj_red) +
    geom_text(aes(label = sprintf("beta = %.3f", estimate)),
              vjust = -1.3, size = 3.5, fontface = "bold") +
    annotate("text", x = stage1_coef$conf.high * 1.05, y = 1,
             label = sprintf("F = %.1f", as.numeric(stage1_ftest$Ftest)),
             size = 3.5, color = wsj_navy, fontface = "bold", hjust = 0) +
    labs(title = "Stage 1: Child BMI Is a Strong Instrument",
         subtitle = "Unit increase in child BMI predicts unit increase in parent BMI",
         x = "Stage 1 coefficient", y = NULL) +
    theme_wsj(base_size = 11) +
    theme(axis.title.x = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

The F-statistic on the excluded instrument is
`r sprintf("%.1f", as.numeric(stage1_ftest$Ftest))`,
`r if(as.numeric(stage1_ftest$Ftest) > 10) "exceeding" else "below"` the conventional
threshold of 10 for a strong instrument.

### Stage 2: IV-Adjusted Spending Model

```{r stage2}
#| cache: true
data_iv <- data_iv |>
    mutate(resid_stage1 = residuals(stage1_model))

svy_iv <- data_iv |>
    as_survey_design(ids = variance_psu, strata = variance_stratum,
                     weights = person_weight, nest = TRUE)

model_2sri <- survey::svyglm(
    total_expenditure ~ bmi + resid_stage1 +
        age_group + sex + race_eth + education + marital_status + region +
        family_size + proxy_reported + child_age_months + as.factor(survey_year) +
        diabetes + hypertension + hyperlipidemia + mental_health +
        pulmonary_disease + arthritis + back_problems + heart_disease +
        cerebrovascular_disease + asthma,
    design = svy_iv, family = Gamma(link = "log")
)
```

```{r stage2-bmi-coef}
iv_bmi_coef <- tibble(
    term = names(coef(model_2sri)), estimate = coef(model_2sri),
    conf.low = confint(model_2sri)[, 1], conf.high = confint(model_2sri)[, 2]
) |>
    filter(term == "bmi") |>
    mutate(across(c(estimate, conf.low, conf.high), exp))

ggplot(iv_bmi_coef, aes(x = estimate, y = "2SRI (IV-Adjusted)")) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "gray50", linewidth = 0.7) +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                   height = 0.15, linewidth = 1.1, color = "gray40") +
    geom_point(size = 5, color = wsj_red) +
    geom_text(aes(label = sprintf("x%.4f", estimate)),
              vjust = -1.3, size = 3.5, fontface = "bold") +
    labs(title = "2SRI: Spending Multiplier Per Unit BMI Increase",
         subtitle = "Exponentiated continuous BMI coefficient | Gamma GLM, IV-adjusted",
         x = "Spending ratio per unit BMI (reference = 1.0)", y = NULL) +
    theme_wsj(base_size = 11) +
    theme(axis.title.x = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

### BMI Reduction Simulation

```{r simulation}
#| cache: true
bmi_pcts <- c(0.05, 0.10, 0.15, 0.20, 0.25)
baseline_pred_iv <- as.numeric(predict(model_2sri, newdata = data_iv, type = "response"))

sim_results <- map(bmi_pcts, \(pct) {
    sim_pred <- as.numeric(predict(model_2sri,
                                   newdata = data_iv |> mutate(bmi = bmi * (1 - pct)),
                                   type = "response"))
    tibble(
        bmi_reduction_pct = pct * 100,
        mean_saving_pp    = weighted.mean(baseline_pred_iv - sim_pred,
                                          w = data_iv$person_weight),
        total_saving_bil  = sum((baseline_pred_iv - sim_pred) *
                                    data_iv$person_weight) / 1e9
    )
}) |> list_rbind()
```

```{r plot-sim}
#| layout-ncol: 2
ggplot(sim_results, aes(x = bmi_reduction_pct, y = mean_saving_pp)) +
    geom_line(color = wsj_navy, linewidth = 1.2) +
    geom_point(size = 4, color = wsj_red) +
    geom_text(aes(label = dollar(round(mean_saving_pp, -1))),
              vjust = -0.9, size = 3.5, fontface = "bold") +
    scale_x_continuous(breaks = bmi_pcts * 100, labels = \(x) paste0(x, "%")) +
    scale_y_continuous(labels = dollar_format(), expand = expansion(mult = c(0.05, 0.2))) +
    labs(title = "Greater BMI Reduction Yields Greater Savings",
         subtitle = "Per-person annual savings | 2SRI IV model",
         x = "BMI reduction", y = "Mean savings per person") +
    theme_wsj(base_size = 11) +
    theme(axis.title = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))

ggplot(sim_results, aes(x = factor(bmi_reduction_pct), y = total_saving_bil)) +
    geom_col(fill = wsj_navy, width = 0.65) +
    geom_text(aes(label = sprintf("$%.1fB", total_saving_bil)),
              vjust = -0.6, size = 4, fontface = "bold") +
    scale_x_discrete(labels = \(x) paste0(x, "%")) +
    scale_y_continuous(labels = \(x) sprintf("$%.0fB", x),
                       expand = expansion(mult = c(0, 0.2))) +
    labs(title = "Estimated Total Medicare Savings",
         subtitle = "Population-weighted | 2SRI IV model",
         x = "BMI reduction", y = "Total savings (billions)") +
    theme_wsj(base_size = 11) +
    theme(axis.title = element_text(size = 9),
          plot.subtitle = element_text(size = 9, color = "gray40"))
```

---

## Comparison of Analyses

### Side-by-Side Results

```{r comparison-data}
comparison_savings <- bind_rows(
    savings_data |>
        mutate(
            method   = "My Analysis\n(Naive GLM)",
            scenario = case_match(as.character(bmi_category),
                "obese_class_i"   ~ "Class I\n-> Overweight",
                "obese_class_ii"  ~ "Class II\n-> Class I",
                "obese_class_iii" ~ "Class III\n-> Class II"),
            savings_pp  = mean_saving_pp,
            savings_tot = total_saving / 1e9
        ) |> select(method, scenario, savings_pp, savings_tot),
    sim_results |>
        mutate(
            method      = "2SRI Replication\n(IV-Adjusted)",
            scenario    = paste0(bmi_reduction_pct, "%"),
            savings_pp  = mean_saving_pp,
            savings_tot = total_saving_bil
        ) |> select(method, scenario, savings_pp, savings_tot)
) |>
    mutate(
        method   = factor(method, levels = c("My Analysis\n(Naive GLM)",
                                             "2SRI Replication\n(IV-Adjusted)")),
        scenario = factor(scenario, levels = unique(scenario))
    )
```

```{r comparison-plots}
#| layout-ncol: 2
ggplot(comparison_savings, aes(x = scenario, y = savings_pp, fill = method)) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(aes(label = dollar(round(savings_pp, -1))),
              vjust = -0.6, size = 3.2, fontface = "bold") +
    scale_fill_manual(values = c("My Analysis\n(Naive GLM)" = wsj_navy,
                                 "2SRI Replication\n(IV-Adjusted)" = wsj_red)) +
    scale_y_continuous(labels = dollar_format(), expand = expansion(mult = c(0, 0.22))) +
    facet_wrap(~ method, scales = "free_x", nrow = 1) +
    labs(title = "Per-Person Savings: Two Methods",
         x = NULL, y = "Mean annual savings per person") +
    theme_wsj(base_size = 10) +
    theme(axis.title.y = element_text(size = 9), strip.text = element_text(face = "bold"))

ggplot(comparison_savings, aes(x = scenario, y = savings_tot, fill = method)) +
    geom_col(width = 0.65, show.legend = FALSE) +
    geom_text(aes(label = sprintf("$%.1fB", savings_tot)),
              vjust = -0.6, size = 3.2, fontface = "bold") +
    scale_fill_manual(values = c("My Analysis\n(Naive GLM)" = wsj_navy,
                                 "2SRI Replication\n(IV-Adjusted)" = wsj_red)) +
    scale_y_continuous(labels = \(x) sprintf("$%.0fB", x),
                       expand = expansion(mult = c(0, 0.22))) +
    facet_wrap(~ method, scales = "free_x", nrow = 1) +
    labs(title = "Total Medicare Savings: Two Methods",
         x = NULL, y = "Total annual savings (billions)") +
    theme_wsj(base_size = 10) +
    theme(axis.title.y = element_text(size = 9), strip.text = element_text(face = "bold"))
```

### Methodological Comparison

```{r method-table}
#| echo: false
tibble(
    Feature             = c("BMI parameterization", "Endogeneity treatment",
                             "Sample", "Controls", "Savings scenario",
                             "Key output"),
    `My Analysis`       = c("Categorical (5 classes)", "None (observational)",
                             "Full analytic sample", "Age, sex, race, ethnicity, education,\nmarital status, region, year, conditions",
                             "One BMI class step-down per obese group",
                             "Per-person & total savings by obesity class"),
    `2SRI Replication`  = c("Continuous (unit BMI)", "2SRI with child BMI instrument",
                             "Subset with household child aged 11-45",
                             "As above plus family size, proxy status,\nchild age in months",
                             "5 / 10 / 15 / 20 / 25% BMI reduction",
                             "Per-person & total savings by % reduction")
) |> knitr::kable()
```

### Discussion

The two analyses answer related but distinct questions and should be interpreted accordingly.

**My Analysis (Naive GLM)** estimates the *conditional association* between BMI class and
annual healthcare spending among Medicare beneficiaries with at least one chronic condition.
After adjusting for demographics and comorbidities, each successive obesity class is
associated with meaningfully higher spending relative to the normal/underweight reference.
The counterfactual projection — stepping every obese beneficiary down one BMI class — is an
accounting exercise: it quantifies what aggregate spending *would have been* if spending
patterns of lower-BMI beneficiaries applied to the current obese population. It is not a
causal estimate.

The naive approach overstates the causal effect of BMI reduction for two reasons. First,
**unmeasured confounding**: individuals with higher BMI may differ from lower-BMI individuals
in ways that independently drive spending — genetic predisposition, access to care, health
behaviors — and the regression cannot fully account for these. Second, **reverse causality**
is possible: poor health may cause both weight gain and high spending, making the observed
association larger than the effect of weight change alone. As a result, naive GLM estimates
should be regarded as an upper bound on the spending impact of obesity reduction programs.

**The 2SRI Replication** addresses endogeneity by exploiting the genetic link between
parent and child BMI. Because child BMI is determined largely by genetics inherited from the
parent — not by the parent's healthcare utilization — it provides an exogenous shift in
parental BMI. The IV estimate captures the *causal* spending effect of BMI, net of
confounders. This is conceptually the right estimand for evaluating the financial impact of
a weight-loss intervention: it reflects the spending change attributable specifically to a
reduction in adiposity rather than to correlated health behaviors or unmeasured risk factors.

**Comparing savings magnitudes** requires care because the two analyses use different units.
The naive model projects savings from a *discrete class transition* (which implies different
BMI reductions depending on where within a class a person sits), while the IV model projects
savings from a *fixed percentage reduction* applied to continuous BMI. A rough comparison:
the naive model's Class I → Overweight transition corresponds to a reduction of roughly
5–15% of BMI for most Class I beneficiaries, and the per-person savings from the naive model
for that group can be read against the IV model's 5–15% reduction estimates.

Where the two methods agree, confidence in the finding is higher. Where they diverge, the
IV estimate is the more defensible figure for policy purposes — particularly for programs
that target BMI reduction through diet, exercise, or pharmacotherapy.

**Limitations of the IV approach** are also worth noting. The instrument restricts the
sample to the subset of Medicare beneficiaries with a household child aged 11–45, which
is a selected group: older Medicare beneficiaries (85+) are unlikely to have children in
that age range, and the instrument sample may skew toward younger or healthier enrollees.
The IV estimate is therefore a Local Average Treatment Effect (LATE) — the effect for
individuals whose BMI is influenced by the genetic endowment transmitted to their children,
which may not generalize to the full Medicare population. Finally, the exclusion restriction
— that child BMI affects parental spending only through parental BMI — could be violated if
shared household health behaviors independently drive both child BMI and parental spending,
though prior research has found limited evidence of such household effects on weight.

---

## Conclusions

Both analyses consistently show that obesity is associated with substantially higher
healthcare spending among Medicare beneficiaries, and that greater BMI reduction implies
greater savings. Hypertension, diabetes, and heart disease show the strongest gradients
across BMI categories in the descriptive data. The naive Gamma GLM provides an intuitive,
accessible summary of the spending gradient and a clear upper-bound savings projection.
The 2SRI IV replication provides a more causally defensible estimate, accounting for the
endogeneity inherent in observational BMI data.

For policy applications — such as estimating the Medicare savings from GLP-1 coverage or
structured weight-loss programs — the IV-based estimates are preferred. For descriptive
and planning purposes, the naive model's categorical framework may be more practical.
Together, the two approaches bracket the likely range of savings attributable to obesity
reduction in this population.

---

*Analysis conducted in R. Data: MEPS-HC 2020 & 2022 (AHRQ). Survey weights applied
throughout via the `srvyr` package. All models account for MEPS stratified cluster
sampling design.*
